\section{Conclusion}

In this project, methods for creating circuits without clocks have
been researched, and a clockless AES-module have been implemented to a
netlist comprising of primitive gates, using the Balsa language and
Teak synthesis system. Well founded theoretical foundations was found,
but the found workflows for implementing clockless circuits, while
able to implement circuits, was not found to be as stable and complete
as their clocked counterparts. Clockless circuits have special
properties, and should be considered when the goals are low power, low
emi, secure crypto implementations and robustness to operating
conditions.

\subsection{Further work}

A clockless AES module was developed in this project, with the purpose
of testing tools for clockless development. While testing have been
implemented for the Tangram tool flow, the alternatives employed in
this project lacks support for this kind of testing. It should be
possible to implement testing both in Balsa and Teak.

Furthermore, technology-mapping and place \& route need further
research to expolore the clockless methodology. Power estimation is
also important to assess how design choices affect power consumption

With Teak, new possibilites for automatic optimization opens up, and
is clearly an area for further research.

\begin{quotation}
  ``I regard programs as specific instances of mechanisms, and I wanted
  to express, at least once, my strong feeling that many of my
  considerations concerning software are, mutatis mutandis, just as
  relevant for hardware design''

  --Edsger W. Dijkstra
\end{quotation}

